%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter*{Abstract}
\thispagestyle{plain}
\addcontentsline{toc}{chapter}{Abstract}

The COVID-19 pandemic moved education online overnight, and with that shift came a problem technology alone did not solve. In a physical classroom, disengagement has visible signals: a student who stopped writing, one staring past the screen, one who has gone quiet. Those signals disappear on a video call. Existing automated detection systems try to recover them but lean on deep learning pipelines that need hardware most schools do not have, and nearly all reduce engagement to a flat binary that ignores how gradually attention fades.

This study uses the DAiSEE benchmark, a standard collection for in-the-wild student engagement recognition covering four affective states: boredom, engagement, confusion, and frustration, each classified at three ordinal levels: Low, Medium, and High. Per frame, 78 features are extracted through computer vision covering facial landmarks, head pose, eye gaze, and action unit intensities, then passed into an XGBoost classifier. Beyond predictions, XGBoost outputs feature important scores that identify which facial cues most influence each decision, making the model transparent in a way neural networks typically are not.

The model achieves a mean accuracy of 65.26\% and a weighted F1-score of 59.62\%, with frustration reaching 80.25\% per-class accuracy. Macro F1 sits at 0.397, pulled down by DAiSEE's class imbalance and the difficulty of separating Mid-level states, a challenge deep learning baselines on this dataset also struggle with. Feature importance analysis reveals that AU-based features account for 60\% of the top-20 ranked features, confirming that facial muscle activity carries more predictive signal than head pose or gaze alone. The full pipeline runs without GPU acceleration, making it deployable on standard classroom hardware.

This work contributes toward SDG4 Quality Education by giving educators a practical, explainable attention detection tool for data-driven teaching in resource-constrained online learning environments where disengagement often goes unnoticed.

% Brief context and background (1-2 sentences)
% This research addresses the challenge of real-time attention detection in online and hybrid learning environments. With the increasing importance of remote education, there is a critical need for automated, scalable, and unbiased methods to monitor student engagement during digital instruction.

% Research objectives and methods (2-3 sentences)
% The primary objective of this research is to develop a lightweight deep learning system capable of detecting multiple affective states indicative of learner attention on resource-constrained devices. To achieve this, we propose a spatio-temporal neural architecture that leverages a TimeDistributed MobileNetV3 backbone integrated with Long Short-Term Memory (LSTM) networks for temporal sequence modeling. The methodology involves training on the DAiSEE dataset using a 70/15/15 train/validation/test split, with multiple independent binary classification heads predicting four key dimensions - boredom, engagement, confusion, and frustration, optimized using the Adam optimizer and binary cross-entropy loss.

% Key contributions and results (2-3 sentences)
% The main contributions of this work include: (1) a computationally efficient spatio-temporal architecture suitable for deployment on low-end devices such as the Raspberry Pi; (2) simultaneous, multi-dimensional affective-state classification achieving accuracies of \sout{95.18\%} (engagement), \sout{95.52\%} (frustration), \sout{91.20\%} (confusion), and \sout{78.87\%} (boredom); and (3) improved classification accuracy over existing baseline methods, despite moderate macro-F1 and AUC scores due to dataset class imbalance. Experimental results show that the proposed approach achieves higher accuracy than state-of-the-art methods while maintaining real-time inference on resource-limited hardware.

% Implications and applications (1-2 sentences)
% These findings have significant implications for adaptive learning systems and automated pedagogical interventions and provide a foundation for scalable student-attention monitoring in large-scale digital classrooms. The proposed framework is applicable to real-world educational technology platforms that require immediate detection of disengagement to enable timely instructional support.

\noindent\textbf{Keywords:} \textit{
Student Engagement Detection; Affective State Classification; Feature Extraction; Extreme Gradient Boosting (XGBoost); Computer Vision
}

\clearpage
